---
title: "NHANES 2021–2023: Dietary Fiber and Depression Severity (Ordinal PHQ-9)"
subtitle: "Survey-Weighted Ordinal Logistic Regression"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 150
)
library(MASS)
library(haven)
library(dplyr)
library(tidyr)
library(survey)
library(splines)
library(ggplot2)
library(knitr)
library(DiagrammeR)

# Ensure dplyr::select wins over MASS::select (loaded by survey)
select <- dplyr::select

options(survey.lonely.psu = "adjust")
set.seed(42)

# Helper: extract OR, 95% CI, and p-value from svyglm.
# Uses Wald z-test (coef + vcov), which always works with quasibinomial.
or_ci_p <- function(fit, coef_name) {
  b  <- unname(coef(fit)[coef_name])
  se <- sqrt(unname(vcov(fit)[coef_name, coef_name]))
  z  <- b / se
  list(
    OR    = exp(b),
    ci_lo = exp(b - qnorm(0.975) * se),
    ci_hi = exp(b + qnorm(0.975) * se),
    p     = 2 * pnorm(-abs(z)),
    se    = se
  )
}

# Helper: extract OR, 95% CI, and p-value from svyolr.
# svyolr stores regression coefficients in $coefficients and
# intercepts in $zeta; vcov covers both, but coefficient names
# are accessible via the same indexing.
or_ci_p_olr <- function(fit, coef_name) {
  b  <- fit$coefficients[coef_name]
  se <- sqrt(diag(vcov(fit)))[coef_name]
  z  <- b / se
  list(
    OR    = exp(b),
    ci_lo = exp(b - qnorm(0.975) * se),
    ci_hi = exp(b + qnorm(0.975) * se),
    p     = 2 * pnorm(-abs(z)),
    se    = se
  )
}

# Format OR (95% CI) as string
fmt_or <- function(r) sprintf("%.2f (%.2f-%.2f)", r$OR, r$ci_lo, r$ci_hi)
fmt_p  <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
```

## Hypotheses

- **H1** -- Higher dietary fiber intake is associated with lower odds of
  greater depression severity (5-level ordinal PHQ-9), both overall and
  by fiber source.
- **H2** -- The fiber--depression relationship follows an inverted-L
  dose-response curve, with diminishing returns beyond a threshold.
- **H3** -- Fiber correlates most strongly with motivational/psychomotor
  PHQ-9 symptoms, least with suicidal ideation.

**Approach.** Survey-weighted ordinal logistic regression (`svyolr`) with
progressive covariate adjustment. The outcome is a 5-level ordered factor
of PHQ-9 severity (None / Mild / Moderate / Moderately Severe / Severe)
matching Varre et al. (2026). ORs represent the odds of being in a higher
(more severe) depression category per 1 SD increase in the exposure.
All continuous predictors are centered and scaled (mean = 0, SD = 1) for
numerical stability, following Varre et al. (2026). The Willett residual
energy-adjusted version is available for sensitivity analysis.


# 1. Data Acquisition

```{r data-download}
DATA_DIR <- ".data"
NHANES_BASE <- "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles"
FILENAMES <- c("DEMO_L", "DR1TOT_L", "DR2TOT_L", "DR1IFF_L", "DR2IFF_L",
               "DPQ_L", "BMX_L", "ALQ_L", "SMQ_L", "PAQ_L")

dir.create(DATA_DIR, showWarnings = FALSE)

for (fname in FILENAMES) {
  fpath <- file.path(DATA_DIR, paste0(fname, ".xpt"))
  if (!file.exists(fpath)) {
    url <- paste0(NHANES_BASE, "/", fname, ".xpt")
    download.file(url, fpath, mode = "wb", quiet = TRUE)
  }
}

dfs <- setNames(lapply(FILENAMES, function(fname) {
  read_xpt(file.path(DATA_DIR, paste0(fname, ".xpt")))
}), FILENAMES)

file_info <- data.frame(
  File = FILENAMES,
  Rows = sapply(dfs, nrow),
  Cols = sapply(dfs, ncol)
)
kable(file_info, row.names = FALSE)
```


# 2. Cohort Construction

```{r cohort}
# Adults >= 18
n_demo_all <- nrow(dfs$DEMO_L)
demo <- dfs$DEMO_L %>%
  select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3, INDFMPIR,
         DMDEDUC2, SDMVPSU, SDMVSTRA, WTMEC2YR) %>%
  filter(RIDAGEYR >= 18)
n_adults <- nrow(demo)

# Two reliable 24-h dietary recalls
dr1 <- dfs$DR1TOT_L
dr2 <- dfs$DR2TOT_L
reliable <- intersect(
  dr1$SEQN[dr1$DR1DRSTZ == 1],
  dr2$SEQN[dr2$DR2DRSTZ == 1]
)

# Fiber + energy (2-day average) + dietary weight
nutr <- dr1 %>%
  filter(SEQN %in% reliable) %>%
  select(SEQN, DR1TFIBE, DR1TKCAL, WTDR2D) %>%
  inner_join(
    dr2 %>% filter(SEQN %in% reliable) %>% select(SEQN, DR2TFIBE, DR2TKCAL),
    by = "SEQN"
  ) %>%
  mutate(
    fiber_total = (DR1TFIBE + DR2TFIBE) / 2,
    energy_kcal = (DR1TKCAL + DR2TKCAL) / 2
  )

# Fiber by source (USDA food codes: 42/43=nuts/seeds, 5=grains, 6=fruits, 7/41=vegetables)
fiber_by_source <- function(iff, fdcd_col, fibe_col) {
  iff %>%
    mutate(
      code = as.character(.data[[fdcd_col]]),
      src = case_when(
        startsWith(code, "42") | startsWith(code, "43") ~ "nuts_seeds",
        startsWith(code, "5") ~ "grains",
        startsWith(code, "6") ~ "fruits",
        startsWith(code, "7") | startsWith(code, "41") ~ "vegetables",
        TRUE ~ "other"
      )
    ) %>%
    group_by(SEQN, src) %>%
    summarise(fiber = sum(.data[[fibe_col]], na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = src, values_from = fiber,
                values_fill = 0, names_prefix = "fiber_")
}

fb1 <- fiber_by_source(dfs$DR1IFF_L, "DR1IFDCD", "DR1IFIBE")
fb2 <- fiber_by_source(dfs$DR2IFF_L, "DR2IFDCD", "DR2IFIBE")

src_cols <- c("fiber_fruits", "fiber_grains", "fiber_nuts_seeds",
              "fiber_other", "fiber_vegetables")
for (col in src_cols) {
  if (!(col %in% names(fb1))) fb1[[col]] <- 0
  if (!(col %in% names(fb2))) fb2[[col]] <- 0
}

fiber_src <- full_join(
  fb1 %>% select(SEQN, all_of(src_cols)),
  fb2 %>% select(SEQN, all_of(src_cols)),
  by = "SEQN", suffix = c("_1", "_2")
)
for (col in src_cols) {
  c1 <- paste0(col, "_1"); c2 <- paste0(col, "_2")
  fiber_src[[col]] <- rowMeans(cbind(
    replace(fiber_src[[c1]], is.na(fiber_src[[c1]]), 0),
    replace(fiber_src[[c2]], is.na(fiber_src[[c2]]), 0)
  ))
}
fiber_src <- fiber_src %>% select(SEQN, all_of(src_cols))

# PHQ-9
dpq_items <- paste0("DPQ0", 1:9, "0")
phq <- dfs$DPQ_L %>% select(SEQN, all_of(dpq_items))
for (item in dpq_items) {
  phq[[item]] <- round(phq[[item]])
  phq[[item]][phq[[item]] > 3] <- NA
}
phq$phq9_total <- rowSums(phq[dpq_items])
phq$phq9_total[rowSums(is.na(phq[dpq_items])) > 0] <- NA

# Covariates
bmi_df <- dfs$BMX_L %>% select(SEQN, BMXBMI)

smq <- dfs$SMQ_L %>%
  select(SEQN, SMQ020, SMQ040) %>%
  mutate(smoking = case_when(
    SMQ020 == 2 ~ 0,
    SMQ020 == 1 & SMQ040 == 3 ~ 1,
    SMQ020 == 1 & SMQ040 %in% c(1, 2) ~ 2
  )) %>%
  select(SEQN, smoking)

alq <- dfs$ALQ_L %>%
  select(SEQN, ALQ111, ALQ121) %>%
  mutate(alcohol = case_when(
    ALQ111 %in% c(7, 9)           ~ NA_integer_,  # refused / don't know
    ALQ111 == 2                    ~ 0L,            # never drank
    ALQ121 == 0                    ~ 0L,            # not in past year
    ALQ121 %in% c(7777, 9999)     ~ NA_integer_,   # refused / don't know
    ALQ121 >= 1                    ~ 1L             # any current use
  )) %>%
  select(SEQN, alcohol)

freq_to_weekly <- function(freq, unit) {
  invalid <- freq %in% c(7777, 9999) | is.na(freq)
  wk <- case_when(
    invalid        ~ NA_real_,
    freq == 0      ~ 0,
    unit == "D"    ~ freq * 7,
    unit == "W"    ~ freq,
    unit == "M"    ~ freq / 4.345,
    unit == "Y"    ~ freq / 52,
    TRUE           ~ NA_real_
  )
  wk
}

paq <- dfs$PAQ_L %>%
  select(SEQN, PAD790Q, PAD790U, PAD800, PAD810Q, PAD810U, PAD820) %>%
  mutate(
    mod_freq_wk = freq_to_weekly(PAD790Q, PAD790U),
    vig_freq_wk = freq_to_weekly(PAD810Q, PAD810U),
    mod_min = ifelse(PAD800 %in% c(7777, 9999), NA_real_, PAD800),
    vig_min = ifelse(PAD820 %in% c(7777, 9999), NA_real_, PAD820),
    # Weekly minutes (moderate-equivalent: vigorous counts 2x per WHO)
    mod_wk_min = ifelse(is.na(mod_freq_wk) | is.na(mod_min), 0, mod_freq_wk * mod_min),
    vig_wk_min = ifelse(is.na(vig_freq_wk) | is.na(vig_min), 0, vig_freq_wk * vig_min),
    meq_min_wk = mod_wk_min + 2 * vig_wk_min,
    # WHO guideline: >= 150 moderate-equivalent minutes/week
    active = case_when(
      PAD790Q %in% c(7777, 9999) & PAD810Q %in% c(7777, 9999) ~ NA_integer_,
      TRUE ~ as.integer(meq_min_wk >= 150)
    )
  ) %>%
  select(SEQN, active)

# Merge (step-by-step for cohort flow tracking)
df <- demo %>% filter(SEQN %in% reliable)
n_reliable <- nrow(df)

df <- df %>%
  inner_join(nutr %>% select(SEQN, fiber_total, energy_kcal, WTDR2D), by = "SEQN")
n_diet <- nrow(df)

df <- df %>%
  left_join(fiber_src, by = "SEQN") %>%
  inner_join(phq, by = "SEQN")
n_phq <- nrow(df)

df <- df %>%
  left_join(bmi_df, by = "SEQN") %>%
  left_join(smq, by = "SEQN") %>%
  left_join(alq, by = "SEQN") %>%
  left_join(paq, by = "SEQN")

for (s in src_cols) df[[s]][is.na(df[[s]])] <- 0

df <- df %>%
  rename(age = RIDAGEYR, bmi = BMXBMI,
         race = RIDRETH3, psu = SDMVPSU, stratum = SDMVSTRA) %>%
  mutate(
    female    = as.integer(RIAGENDR == 2),
    depressed = as.integer(phq9_total >= 10),
    education = ifelse(DMDEDUC2 %in% c(7, 9), NA, DMDEDUC2),
    poverty   = as.integer(INDFMPIR <= 1),
    race      = factor(race),
    education = factor(education),
    smoking   = factor(smoking)
  )

# 5-level ordinal depression severity (Varre et al. 2026)
df$dep_severity <- cut(df$phq9_total,
                       breaks = c(-1, 4, 9, 14, 19, 27),
                       labels = c("None", "Mild", "Moderate", "Mod.Severe", "Severe"))
df$dep_severity <- ordered(df$dep_severity)

# Energy-adjusted fiber (Willett residual method, 1997)
valid_idx <- !is.na(df$fiber_total) & !is.na(df$energy_kcal)
fit_energy <- lm(fiber_total ~ energy_kcal, data = df, subset = valid_idx)
df$fiber_adj <- NA_real_
df$fiber_adj[valid_idx] <- residuals(fit_energy) + mean(df$fiber_total[valid_idx])
df$fiber_adj_5g <- df$fiber_adj / 5

# Raw fiber / 5, then centered+scaled below (ORs are per 1 SD)
df$fiber_5g <- df$fiber_total / 5

# Final analytic sample
df <- df %>% filter(!is.na(phq9_total), !is.na(fiber_total), !is.na(WTDR2D))
n_analytic <- nrow(df)

# Center and scale continuous predictors for svyolr numerical stability
# (following Varre et al. 2026). Keep unscaled copies for descriptive stats.
for (v in c("fiber_5g", "age", "bmi", "energy_kcal")) {
  df[[paste0(v, "_raw")]] <- df[[v]]
  df[[v]] <- as.numeric(scale(df[[v]]))
}
# Also scale fiber_adj_5g
df$fiber_adj_5g_raw <- df$fiber_adj_5g
df$fiber_adj_5g <- as.numeric(scale(df$fiber_adj_5g))

# Count for fully adjusted models (complete covariates)
fully_adj_vars <- c("dep_severity", "fiber_5g", "age", "female", "race",
                    "education", "poverty", "bmi", "smoking", "alcohol", "active",
                    "energy_kcal", "WTDR2D", "psu", "stratum")
n_fully_adj <- sum(complete.cases(df[fully_adj_vars]))
```

```{r cohort-flow, fig.height=7}
fmt <- function(n) format(n, big.mark = ",")

flow_diagram <- grViz(sprintf('
digraph cohort {
  graph [rankdir=TB, nodesep=0.5, ranksep=0.35]
  node [shape=box, style="filled,rounded", fontname=Helvetica, fontsize=10,
        margin="0.2,0.1"]
  edge [arrowsize=0.8]

  node [fillcolor="#E3F2FD", width=3.2]
  n0 [label="NHANES 2021-2023 participants\\nN = %s"]
  n1 [label="Adults (age >= 18)\\nN = %s"]
  n2 [label="Two reliable 24-h dietary recalls\\nN = %s"]
  n3 [label="Complete fiber & energy data\\nN = %s"]
  n4 [label="PHQ-9 administered\\nN = %s"]
  n5 [label="Analytic sample\\nN = %s", fillcolor="#BBDEFB"]
  n6 [label="Fully adjusted models\\nN = %s", fillcolor="#90CAF9"]

  node [fillcolor="#FFF9C4", width=2.4, fontsize=9]
  e1 [label="Excluded (n = %s)\\nAge < 18"]
  e2 [label="Excluded (n = %s)\\nMissing/unreliable\\ndietary recall"]
  e3 [label="Excluded (n = %s)\\nMissing fiber or\\nenergy values"]
  e4 [label="Excluded (n = %s)\\nNo PHQ-9 data"]
  e5 [label="Excluded (n = %s)\\nIncomplete PHQ-9\\nor missing weights"]
  e6 [label="Excluded (n = %s)\\nMissing covariates\\n(education, BMI, smoking,\\nalcohol, physical activity)"]

  n0 -> n1 -> n2 -> n3 -> n4 -> n5 -> n6
  n0 -> e1
  n1 -> e2
  n2 -> e3
  n3 -> e4
  n4 -> e5
  n5 -> e6

  {rank=same; n0; e1}
  {rank=same; n1; e2}
  {rank=same; n2; e3}
  {rank=same; n3; e4}
  {rank=same; n4; e5}
  {rank=same; n5; e6}
}',
  fmt(n_demo_all),
  fmt(n_adults),
  fmt(n_reliable),
  fmt(n_diet),
  fmt(n_phq),
  fmt(n_analytic),
  fmt(n_fully_adj),
  fmt(n_demo_all - n_adults),
  fmt(n_adults - n_reliable),
  fmt(n_reliable - n_diet),
  fmt(n_diet - n_phq),
  fmt(n_phq - n_analytic),
  fmt(n_analytic - n_fully_adj)
))

flow_diagram

# Export to PDF
svg_txt <- DiagrammeRsvg::export_svg(flow_diagram)
rsvg::rsvg_pdf(charToRaw(svg_txt), "cohort_flow_ordinal.pdf")
```

**Analytic sample:** N = `r format(n_analytic, big.mark = ",")` adults.
Depression severity distribution:
`r paste(levels(df$dep_severity), sapply(levels(df$dep_severity), function(lv) sprintf("%.1f%%", 100 * mean(df$dep_severity == lv, na.rm = TRUE))), sep = ": ", collapse = "; ")`.
Fiber intake: `r sprintf("%.1f +/- %.1f", mean(df$fiber_total), sd(df$fiber_total))` g/day.

```{r survey-design}
des <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                 nest = TRUE, data = df)
```


# 3. Descriptive Statistics

```{r table1}
severity_levels <- levels(df$dep_severity)

# Label maps for categorical variables
race_labels <- c("1" = "Mexican American", "2" = "Other Hispanic",
                 "3" = "Non-Hispanic White", "4" = "Non-Hispanic Black",
                 "6" = "Non-Hispanic Asian", "7" = "Other/Multi-Racial")
edu_labels  <- c("1" = "Less than 9th grade", "2" = "9-11th grade",
                 "3" = "High school grad/GED", "4" = "Some college/AA",
                 "5" = "College graduate+")
smoke_labels <- c("0" = "Never", "1" = "Former", "2" = "Current")

# --- P-value helpers --------------------------------------------------------
p_wald <- function(var) {
  f <- as.formula(paste(var, "~ dep_severity"))
  m <- svyglm(f, design = des)
  p <- regTermTest(m, "dep_severity")$p
  fmt_p(p)
}

p_chisq <- function(var) {
  f <- as.formula(paste("~", var, "+ dep_severity"))
  p <- svychisq(f, design = des)$p.value
  fmt_p(p)
}

# --- Row helpers ------------------------------------------------------------
make_row_ord <- function(var, label) {
  vals <- lapply(severity_levels, function(lv) {
    x <- df[[var]][df$dep_severity == lv]
    sprintf("%.1f +/- %.1f", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
  })
  row <- data.frame(Variable = label, stringsAsFactors = FALSE)
  for (i in seq_along(severity_levels)) row[[severity_levels[i]]] <- vals[[i]]
  row$P <- p_wald(var)
  row
}

make_pct_row_ord <- function(var, label) {
  vals <- lapply(severity_levels, function(lv) {
    x <- df[[var]][df$dep_severity == lv]
    sprintf("%.1f%%", 100 * mean(x, na.rm = TRUE))
  })
  row <- data.frame(Variable = label, stringsAsFactors = FALSE)
  for (i in seq_along(severity_levels)) row[[severity_levels[i]]] <- vals[[i]]
  row$P <- p_chisq(var)
  row
}

make_n_row <- function() {
  vals <- lapply(severity_levels, function(lv) {
    as.character(sum(df$dep_severity == lv, na.rm = TRUE))
  })
  row <- data.frame(Variable = "N", stringsAsFactors = FALSE)
  for (i in seq_along(severity_levels)) row[[severity_levels[i]]] <- vals[[i]]
  row$P <- ""
  row
}

make_cat_rows_ord <- function(var, header_label, level_labels) {
  lvls <- names(level_labels)
  # Header row with p-value
  header <- data.frame(Variable = header_label, stringsAsFactors = FALSE)
  for (sl in severity_levels) header[[sl]] <- ""
  header$P <- p_chisq(var)
  # One sub-row per level
  sub_rows <- lapply(lvls, function(k) {
    row <- data.frame(Variable = paste0("\u00a0\u00a0\u00a0", level_labels[[k]]),
                      stringsAsFactors = FALSE)
    for (sl in severity_levels) {
      x <- df[[var]][df$dep_severity == sl]
      row[[sl]] <- sprintf("%.1f%%", 100 * mean(x == k, na.rm = TRUE))
    }
    row$P <- ""
    row
  })
  bind_rows(header, bind_rows(sub_rows))
}

tab1 <- bind_rows(
  make_n_row(),
  make_row_ord("age_raw", "Age (years)"),
  make_pct_row_ord("female", "Female (%)"),
  make_cat_rows_ord("race", "Race/ethnicity", race_labels),
  make_cat_rows_ord("education", "Education", edu_labels),
  make_pct_row_ord("poverty", "Below poverty line (%)"),
  make_row_ord("bmi_raw", "BMI (kg/m^2)"),
  make_cat_rows_ord("smoking", "Smoking status", smoke_labels),
  make_pct_row_ord("alcohol", "Current drinker (%)"),
  make_pct_row_ord("active", "Physically active (%)"),
  make_row_ord("energy_kcal_raw", "Energy (kcal/day)"),
  make_row_ord("fiber_total", "Fiber (g/day)"),
  make_row_ord("fiber_fruits", "Fiber: fruits (g/day)"),
  make_row_ord("fiber_vegetables", "Fiber: vegetables (g/day)"),
  make_row_ord("fiber_grains", "Fiber: grains (g/day)"),
  make_row_ord("fiber_nuts_seeds", "Fiber: nuts/seeds (g/day)")
)

kable(tab1, align = c("l", rep("r", length(severity_levels)), "r"),
      caption = "Table 1. Sample characteristics by PHQ-9 depression severity category")
```


# 4. H1: Fiber Intake and Depression Severity (Ordinal)

## Collinearity diagnostic

Fiber and energy are correlated. Including both as covariates can inflate
fiber's standard error without changing the point estimate -- classic
collinearity, not confounding.

```{r collinearity}
r_spearman <- cor(df$fiber_total, df$energy_kcal, use = "complete.obs", method = "spearman")

fix_vars <- c("dep_severity", "fiber_5g", "age", "female", "race",
              "poverty", "education", "energy_kcal",
              "WTDR2D", "psu", "stratum")
d_fix <- df %>% filter(complete.cases(pick(all_of(fix_vars))))
des_fix <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                     nest = TRUE, data = d_fix)

fit_no_e <- svyolr(
  dep_severity ~ fiber_5g + age + female + race + poverty + education,
  design = des_fix
)
fit_w_e <- svyolr(
  dep_severity ~ fiber_5g + age + female + race + poverty + education + energy_kcal,
  design = des_fix
)

r_no <- or_ci_p_olr(fit_no_e, "fiber_5g")
r_we <- or_ci_p_olr(fit_w_e,  "fiber_5g")

collin_tab <- data.frame(
  Model = c("Without energy", "With energy"),
  N = nrow(d_fix),
  OR = c(r_no$OR, r_we$OR),
  SE = c(r_no$se, r_we$se),
  p  = c(r_no$p, r_we$p),
  check.names = FALSE
) %>%
  mutate(OR = sprintf("%.3f", OR), SE = sprintf("%.4f", SE), p = sapply(p, fmt_p))

kable(collin_tab,
      caption = sprintf("Fiber-energy Spearman r = %.2f. Same sample (N = %s). Ordinal outcome (svyolr).",
                        r_spearman, format(nrow(d_fix), big.mark = ",")))
```

**OR change:** `r sprintf("%.3f", abs(r_we$OR - r_no$OR))` (negligible -- energy
is not a confounder).
**SE ratio:** `r sprintf("%.2fx", r_we$se / r_no$se)` (collinearity inflates
SE, destroying power).

**Approach:** For the main ordinal models, we use `fiber_5g` (fiber / 5,
then centered and scaled) with energy included as a covariate in the
fully adjusted model, following Varre et al. (2026). The Willett residual
energy-adjusted variable (`fiber_adj_5g`) is available as a sensitivity
check but is not the primary exposure.

## Progressive covariate adjustment

ORs from `svyolr` represent the odds of being in a **higher** (more severe)
depression category per 1 SD increase in fiber intake (1 SD ~ 1.8 g/day
in the `fiber_5g` scale, or ~ 8.8 g/day raw). OR < 1 means fiber is
associated with lower depression severity.

```{r h1-progressive}
model_specs <- list(
  list(name = "Model 1: Crude",
       formula = dep_severity ~ fiber_5g,
       req = c("dep_severity", "fiber_5g")),
  list(name = "Model 2: + Demographics",
       formula = dep_severity ~ fiber_5g + age + female + race,
       req = c("dep_severity", "fiber_5g", "age", "female", "race")),
  list(name = "Model 3: Fully adjusted",
       formula = dep_severity ~ fiber_5g + age + female + race + education +
         poverty + bmi + smoking + alcohol + active + energy_kcal,
       req = c("dep_severity", "fiber_5g", "age", "female", "race",
               "education", "poverty", "bmi", "smoking", "alcohol", "active",
               "energy_kcal"))
)

prog_rows <- lapply(model_specs, function(spec) {
  d <- df %>% filter(complete.cases(pick(all_of(c(spec$req, "WTDR2D", "psu", "stratum")))))
  des_m <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                     nest = TRUE, data = d)
  fit <- svyolr(spec$formula, design = des_m)
  r <- or_ci_p_olr(fit, "fiber_5g")
  data.frame(Model = spec$name, N = nrow(d),
             `OR (95% CI)` = fmt_or(r), P = fmt_p(r$p), check.names = FALSE)
})

prog_tab <- bind_rows(prog_rows)
kable(prog_tab,
      caption = "Ordinal OR per 1 SD fiber increment (svyolr, proportional odds)")
```

```{r h1-progressive-print}
cat("\n=== Progressive Covariate Adjustment (Ordinal, svyolr) ===\n")
for (i in seq_len(nrow(prog_tab))) {
  cat(sprintf("  %s  |  N = %s  |  OR (95%% CI) = %s  |  P = %s\n",
              prog_tab$Model[i], format(prog_tab$N[i], big.mark = ","),
              prog_tab$`OR (95% CI)`[i], prog_tab$P[i]))
}
```

## Quartile analysis (fully adjusted, ordinal)

```{r h1-quartile}
df$fiber_q <- cut(df$fiber_total,
                  breaks = quantile(df$fiber_total, probs = 0:4/4, na.rm = TRUE),
                  include.lowest = TRUE, labels = 1:4)
df$fiber_q <- relevel(factor(df$fiber_q), ref = "1")

q_bounds <- data.frame(
  Quartile = paste0("Q", 1:4),
  Range = sapply(1:4, function(q) {
    sub <- df$fiber_total[df$fiber_q == q]
    sprintf("%.1f-%.1f", min(sub), max(sub))
  }),
  Median = sapply(1:4, function(q) sprintf("%.1f", median(df$fiber_total[df$fiber_q == q])))
)
kable(q_bounds, caption = "Fiber quartile boundaries (g/day)")

req_q <- c("dep_severity", "fiber_q", "age", "female", "race", "education",
           "poverty", "bmi", "smoking", "alcohol", "active", "energy_kcal",
           "WTDR2D", "psu", "stratum")
d_q <- df %>% filter(complete.cases(pick(all_of(req_q))))
des_q <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                   nest = TRUE, data = d_q)

fit_q <- svyolr(
  dep_severity ~ fiber_q + age + female + race + education +
    poverty + bmi + smoking + alcohol + active + energy_kcal,
  design = des_q
)

q_rows <- list(data.frame(Quartile = "Q1 (ref)", N = sum(d_q$fiber_q == 1),
                           `OR (95% CI)` = "1.00", P = "", check.names = FALSE))
for (q in 2:4) {
  key <- paste0("fiber_q", q)
  r <- or_ci_p_olr(fit_q, key)
  q_rows[[q]] <- data.frame(Quartile = paste0("Q", q), N = sum(d_q$fiber_q == q),
                              `OR (95% CI)` = fmt_or(r), P = fmt_p(r$p), check.names = FALSE)
}

# P for trend (use quartile medians as continuous predictor)
d_q$fiber_q_mid <- ave(d_q$fiber_total, d_q$fiber_q, FUN = median)
des_q_trend <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                         nest = TRUE, data = d_q)
fit_trend <- svyolr(
  dep_severity ~ fiber_q_mid + age + female + race + education +
    poverty + bmi + smoking + alcohol + active + energy_kcal,
  design = des_q_trend
)
r_trend <- or_ci_p_olr(fit_trend, "fiber_q_mid")

q_tab <- bind_rows(q_rows)
kable(q_tab,
      caption = sprintf("Quartile analysis (ordinal, fully adjusted, Q1 = reference). P for trend: %s",
                        fmt_p(r_trend$p)))
```

```{r h1-quartile-print}
cat("\n=== Quartile Analysis (Ordinal, svyolr) ===\n")
for (i in seq_len(nrow(q_tab))) {
  cat(sprintf("  %s  |  N = %s  |  OR (95%% CI) = %s  |  P = %s\n",
              q_tab$Quartile[i], format(q_tab$N[i], big.mark = ","),
              q_tab$`OR (95% CI)`[i], q_tab$P[i]))
}
cat(sprintf("  P for trend: %s\n", fmt_p(r_trend$p)))
```

## Fiber by source (fully adjusted, ordinal)

```{r h1-source}
src_rows <- lapply(c("fiber_fruits", "fiber_vegetables", "fiber_grains",
                     "fiber_nuts_seeds"), function(src) {
  d_src <- df
  d_src$src_5g <- as.numeric(scale(d_src[[src]] / 5))
  req_src <- c("dep_severity", "src_5g", "age", "female", "race", "education",
               "poverty", "bmi", "smoking", "alcohol", "active", "energy_kcal",
               "WTDR2D", "psu", "stratum")
  d_src <- d_src %>% filter(complete.cases(pick(all_of(req_src))))
  des_src <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                       nest = TRUE, data = d_src)

  fit_src <- svyolr(
    dep_severity ~ src_5g + age + female + race + education +
      poverty + bmi + smoking + alcohol + active + energy_kcal,
    design = des_src
  )

  r <- or_ci_p_olr(fit_src, "src_5g")
  label <- gsub("fiber_", "", src) %>% gsub("_", "/", .)
  data.frame(Source = label, N = nrow(d_src),
             `OR (95% CI)` = fmt_or(r), P = fmt_p(r$p), check.names = FALSE)
})

src_tab <- bind_rows(src_rows)
kable(src_tab, caption = "Ordinal OR per 1 SD by fiber source (svyolr, fully adjusted)")
```

```{r h1-source-print}
cat("\n=== Fiber by Source (Ordinal, svyolr) ===\n")
for (i in seq_len(nrow(src_tab))) {
  cat(sprintf("  %s  |  N = %s  |  OR (95%% CI) = %s  |  P = %s\n",
              src_tab$Source[i], format(src_tab$N[i], big.mark = ","),
              src_tab$`OR (95% CI)`[i], src_tab$P[i]))
}
```

## H1: Interpretation

ORs from the ordinal model represent the proportional odds of being in a
**higher** depression severity category per 1 SD increase in the
exposure. OR < 1 indicates that higher fiber intake is associated with
lower depression severity.

The ordinal approach preserves the full severity gradient (None through
Severe) rather than collapsing to a binary cutoff at PHQ-9 >= 10. This
increases statistical power, particularly for detecting associations that
operate across the full range of depressive symptoms rather than only at
the clinical threshold.


# 5. H2: Dose-Response (Supplementary, Binary Outcome)

**Note:** The dose-response visualization uses a binary depression outcome
(PHQ-9 >= 10) with `svyglm` and restricted cubic splines. `svyolr` does
not readily support spline-based prediction grids for ordinal outcomes, so
this section is provided as supplementary material. The main results above
use the ordinal model.

```{r h2-models}
req_h2 <- c("depressed", "fiber_total", "age", "female", "race", "education",
             "poverty", "bmi", "smoking", "alcohol", "active", "energy_kcal",
             "WTDR2D", "psu", "stratum")
d_h2 <- df %>%
  filter(complete.cases(pick(all_of(req_h2)))) %>%
  filter(fiber_total <= quantile(fiber_total, 0.97))

des_h2 <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                     nest = TRUE, data = d_h2)

covars <- "+ age + female + race + education + poverty + bmi + smoking + alcohol + active + energy_kcal"

fit_rcs <- svyglm(
  as.formula(paste("depressed ~ ns(fiber_total, df = 3)", covars)),
  design = des_h2, family = quasibinomial()
)

fit_lin <- svyglm(
  as.formula(paste("depressed ~ fiber_total", covars)),
  design = des_h2, family = quasibinomial()
)

# Wald F-test for nonlinearity
rcs_coefs <- names(coef(fit_rcs))
lin_coefs <- names(coef(fit_lin))
nonlin_coefs <- setdiff(rcs_coefs, lin_coefs)

beta_nl <- coef(fit_rcs)[nonlin_coefs]
V_nl    <- vcov(fit_rcs)[nonlin_coefs, nonlin_coefs]
wald_chi2 <- as.numeric(t(beta_nl) %*% solve(V_nl) %*% beta_nl)
ndf <- length(nonlin_coefs)
ddf <- degf(des_h2)
F_nonlin <- wald_chi2 / ndf
p_nonlin <- pf(F_nonlin, ndf, ddf, lower.tail = FALSE)
```

**Nonlinearity test** (Wald F, binary outcome): F = `r sprintf("%.2f", F_nonlin)`,
df~1~ = `r ndf`, df~2~ = `r ddf`,
p = `r fmt_p(p_nonlin)`.

```{r h2-plot, fig.height=5}
get_mode <- function(x) {
  x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

grid <- data.frame(
  fiber_total   = seq(min(d_h2$fiber_total), max(d_h2$fiber_total), length.out = 300),
  age           = median(d_h2$age),
  female        = 0,
  race          = factor(get_mode(d_h2$race), levels = levels(d_h2$race)),
  education     = factor(get_mode(d_h2$education), levels = levels(d_h2$education)),
  poverty       = get_mode(d_h2$poverty),
  bmi           = median(d_h2$bmi, na.rm = TRUE),
  smoking       = factor(get_mode(d_h2$smoking), levels = levels(d_h2$smoking)),
  alcohol       = get_mode(d_h2$alcohol),
  active        = get_mode(d_h2$active),
  energy_kcal   = median(d_h2$energy_kcal, na.rm = TRUE)
)

# OR relative to lowest fiber intake as reference
mm <- model.matrix(delete.response(terms(fit_rcs)), data = grid)
eta <- as.numeric(mm %*% coef(fit_rcs))
se  <- sqrt(rowSums((mm %*% vcov(fit_rcs)) * mm))

ref_idx <- 1  # lowest fiber intake as reference
d_eta <- eta - eta[ref_idx]
d_se  <- sqrt(rowSums(((mm - matrix(mm[ref_idx, ], nrow = nrow(mm),
              ncol = ncol(mm), byrow = TRUE)) %*% vcov(fit_rcs)) *
              (mm - matrix(mm[ref_idx, ], nrow = nrow(mm),
              ncol = ncol(mm), byrow = TRUE))))
grid$OR    <- exp(d_eta)
grid$ci_lo <- exp(d_eta - 1.96 * d_se)
grid$ci_hi <- exp(d_eta + 1.96 * d_se)

mm_lin <- model.matrix(delete.response(terms(fit_lin)), data = grid)
eta_lin <- as.numeric(mm_lin %*% coef(fit_lin))
grid$OR_lin <- exp(eta_lin - eta_lin[ref_idx])

p_dose <- ggplot() +
  geom_ribbon(data = grid, aes(x = fiber_total, ymin = ci_lo, ymax = ci_hi),
              fill = "#1565C0", alpha = 0.15) +
  geom_line(data = grid, aes(x = fiber_total, y = OR),
            color = "#1565C0", linewidth = 1.2) +
  geom_line(data = grid, aes(x = fiber_total, y = OR_lin),
            color = "grey50", linewidth = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = grid$fiber_total[ref_idx],
             linetype = "dotted", color = "grey60", linewidth = 0.4) +
  labs(x = "Fiber intake (g/day)",
       y = "Multivariate-adjusted OR (binary: PHQ-9 >= 10)",
       title = sprintf("Supplementary: Dose-Response (Binary Outcome, N = %s)",
                       format(nrow(d_h2), big.mark = ",")),
       subtitle = sprintf("Reference: lowest intake (%.1f g/day). Binary outcome used for visualization only.",
                          grid$fiber_total[ref_idx])) +
  theme_minimal()
p_dose
ggsave("dose_response_ordinal.pdf", p_dose, width = 9, height = 5)
```



# 6. H3: Fiber and Individual PHQ-9 Symptoms

Each item dichotomized at >= 2 ("more than half the days"). These are
inherently binary outcomes, so `svyglm(..., quasibinomial())` is used
(not ordinal regression).

```{r h3-items}
phq9_labels <- c(
  DPQ010 = "1. Anhedonia",      DPQ020 = "2. Depressed mood",
  DPQ030 = "3. Sleep problems", DPQ040 = "4. Fatigue/energy",
  DPQ050 = "5. Appetite changes", DPQ060 = "6. Feeling of failure",
  DPQ070 = "7. Concentration",  DPQ080 = "8. Psychomotor",
  DPQ090 = "9. Suicidal ideation"
)

for (item in dpq_items) {
  df[[paste0(item, "_bin")]] <- ifelse(df[[item]] >= 2, 1, 0)
  df[[paste0(item, "_bin")]][is.na(df[[item]])] <- NA
}

req_base <- c("fiber_5g", "age", "female", "race", "education",
              "poverty", "bmi", "smoking", "alcohol", "active", "energy_kcal",
              "WTDR2D", "psu", "stratum")

item_results <- lapply(dpq_items, function(item) {
  outcome <- paste0(item, "_bin")
  d_item <- df %>% filter(complete.cases(pick(all_of(c(outcome, req_base)))))
  des_item <- svydesign(ids = ~psu, strata = ~stratum, weights = ~WTDR2D,
                        nest = TRUE, data = d_item)

  f <- as.formula(paste(outcome, "~ fiber_5g + age + female + race + education +",
                        "poverty + bmi + smoking + alcohol + active + energy_kcal"))
  fit_item <- svyglm(f, design = des_item, family = quasibinomial())

  r <- or_ci_p(fit_item, "fiber_5g")
  prev <- mean(d_item[[outcome]], na.rm = TRUE)

  data.frame(item = item, label = phq9_labels[[item]],
             OR = r$OR, ci_lo = r$ci_lo, ci_hi = r$ci_hi,
             p = r$p, prev = prev, stringsAsFactors = FALSE)
}) %>% bind_rows()

item_table <- item_results %>%
  mutate(Prevalence = sprintf("%.1f%%", prev * 100),
         `OR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", OR, ci_lo, ci_hi),
         P = sapply(p, fmt_p)) %>%
  select(Item = label, Prevalence, `OR (95% CI)`, P)

kable(item_table,
      caption = "Survey-weighted OR per 1 SD fiber by PHQ-9 item (fully adjusted, svyglm)")
```

```{r h3-items-print}
cat("\n=== Symptom-Level Analysis (Binary, svyglm) ===\n")
for (i in seq_len(nrow(item_table))) {
  cat(sprintf("  %s  |  Prev = %s  |  OR (95%% CI) = %s  |  P = %s\n",
              item_table$Item[i], item_table$Prevalence[i],
              item_table$`OR (95% CI)`[i], item_table$P[i]))
}
```

```{r h3-forest, fig.height=5}
item_results$label <- factor(item_results$label, levels = rev(item_results$label))
item_results$sig <- factor(
  ifelse(item_results$p < 0.05, "p < 0.05", "n.s."),
  levels = c("p < 0.05", "n.s.")
)

p_forest <- ggplot(item_results, aes(x = OR, y = label, xmin = ci_lo, xmax = ci_hi, color = sig)) +
  geom_errorbarh(height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.5) +
  scale_color_manual(values = c("p < 0.05" = "#E53935", "n.s." = "#9E9E9E"),
                     drop = FALSE) +
  labs(x = "OR per 1 SD fiber (95% CI)", y = NULL, color = NULL,
       title = "Fiber-Symptom Associations by PHQ-9 Item (Fully Adjusted)") +
  theme_minimal() +
  theme(legend.position = "bottom")
p_forest
ggsave("symptom_forest_ordinal.pdf", p_forest, width = 9, height = 5)
```

```{r h3-cicheck}
ci_check <- item_results %>%
  mutate(ci_width = ci_hi - ci_lo,
         flag = ifelse(ci_width < 0.10, "suspect", "")) %>%
  select(Item = label, `CI width` = ci_width, Flag = flag)
kable(ci_check, digits = 3, caption = "CI width check (< 0.10 would indicate artifact)")
```

## H3: Interpretation

Symptom-level results should be interpreted cautiously given the number
of comparisons (9 items tested). Any individual significant associations
(p < 0.05) may reflect chance findings without formal multiple testing
correction.

Biologically, anhedonia (loss of interest or pleasure) and sleep problems
are the most plausible candidates for fiber-related associations via the
gut-brain axis — anhedonia through effects on dopaminergic reward
pathways, sleep through fiber's influence on gut microbiota and
serotonin/melatonin metabolism.

Suicidal ideation has very low prevalence (~1%), leaving insufficient
power to detect an association even if one exists.


# 7. Summary

This analysis uses **survey-weighted ordinal logistic regression** (`svyolr`)
with a 5-level PHQ-9 severity outcome (None / Mild / Moderate / Moderately
Severe / Severe), following Varre et al. (2026). ORs represent the
proportional odds of being in a **higher** (more severe) depression
category per unit increase in the exposure.

The key advantage of the ordinal approach over the binary (PHQ-9 >= 10)
model is that it retains information about the full spectrum of depression
severity rather than collapsing to a single threshold.

### Main findings

1. **Progressive adjustment.** The crude association between fiber and
   depression severity, and its attenuation with SES adjustment, can be
   assessed via the ordinal ORs. The pattern of confounding by
   poverty/education is expected to mirror the binary analysis.

2. **Quartile analysis.** The ordinal quartile model tests whether higher
   fiber quartiles are associated with lower odds of more severe depression,
   with Q1 as reference.

3. **Fiber by source.** Source-specific ordinal ORs identify which fiber
   sources independently predict lower depression severity after full
   covariate adjustment.

4. **Dose-response.** A supplementary binary-outcome dose-response curve
   is provided for visualization. Ordinal models do not easily produce
   continuous dose-response curves because predicted probabilities are
   category-specific.

5. **Symptom-level analysis.** Individual PHQ-9 items are inherently
   binary and are analyzed with `svyglm(..., quasibinomial())` as before.

### Methods notes

- **Ordinal model:** `svyolr()` from the `survey` package fits a
  proportional odds model with survey weights, PSU clustering, and
  stratification. The proportional odds assumption implies that the OR
  for fiber is constant across all severity thresholds.
- **Exposure:** Fiber intake (`fiber_total / 5`), centered and scaled
  to mean = 0, SD = 1 before model fitting. ORs represent a 1 SD
  increase. All continuous predictors (age, BMI, energy) are similarly
  scaled for numerical stability, following Varre et al. (2026).
- **Energy:** Included as a covariate in all fully adjusted models
  (quartile, source, and progressive Model 3).
- **Survey design:** `svydesign()` with PSU clustering, stratification,
  and NHANES 2-day dietary weights (WTDR2D). CIs use Wald z-approximation
  via coefficient estimates and `vcov()`.
- **Covariates (fully adjusted):** age, sex, race/ethnicity, education,
  poverty (PIR <= 1), BMI, smoking (factor: never/former/current),
  alcohol (binary), physical activity (binary: >= 150 MEQ min/week),
  energy (kcal/day).
- **Dose-response (supplementary):** Uses binary outcome with `svyglm`
  and natural splines (df = 3) because `svyolr` does not support
  spline-based prediction grids for ordinal outcomes.

### Limitations

- Cross-sectional design; causal inference is not possible.
- 24-hour dietary recall may not capture habitual intake.
- The proportional odds assumption is not formally tested here; violation
  would mean ORs vary across severity thresholds.
- The dose-response visualization uses a binary outcome, not the ordinal
  model, for technical reasons.
- Binary covariates (poverty, alcohol, physical activity) may provide
  coarser confounding control than continuous alternatives.
